---
title: "Simulating SOC Measurement"
author: "Jake Spertus"
date: "October 31st, 2019"
output: html_notebook
---

```{r knitr setup, message = FALSE, warning = FALSE, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r packages}
library(tidyverse)
library(Matrix)
```

## Goals

This notebook serves to describe and implement a simulation to test the procedures currently used to measure soil organic carbon (SOC) against a known groundtruth. Questions to be answered include:

- With what accuracy can we recover true SOC on average (mean squared error)?
- Are estimates of SOC likely to center on the truth (bias)?
- Are estimates of uncertainty (standard errors) reflecting the true degree of uncertainty in the estimates (coverage of confidence intervals / type 1 error rate)? 


## Empirical Input: The Marin Data

Simulations will be based off of samples from Marin County collected by the Silver Lab at UC Berkeley.

```{r duplicate data}
duplicate_data <- read_csv("mcp_duplicates.csv")
#drop samples that were rerun fewer or more than 2 times
multiple_dups <- names(which(table(duplicate_data$sample_name) != 2))
#also drop runs of atropine standards
duplicate_data <- duplicate_data %>%
  filter(!(sample_name %in% multiple_dups)) %>%
  filter(substr(sample_name, 1, 4) != 'atro') 
```


## Simulating Percent SOC

As a first step, we simulate a "ground truth" (so to speak), which is a 3-dimensional grid on a rectangular prism that represents SOC concentration (in percent) over the plot of land to be studied. The dimensions are `latitude`, `longitude`, and `depth` (in that order). The scale is in meters. 

However, simulating a field of % SOC is not entirely straightforward. In particular we would like to generate numbers that:

- Are bounded between 0 and 1 (0% and 100% SOC).
- Allow for arbitrary correlation between points.
- Have an adjustable mean and variance.

We can achieve these properties using a *Gaussian copula* (see [this post](https://stats.stackexchange.com/questions/37424/how-to-simulate-from-a-gaussian-copula/64988#64988)). Essentially we will simulate values form a multivariate normal distribution and then pass each one through a standard normal CDF to get uniformly distributed, correlated values. To simulate $d$ uniform [0,1] random variables with correlation matrix $P$, we follow these steps:

1) Perform a Cholesky decomposition of $P$ and take $A$ to be the resulting lower triangular matrix. 

2) Generate $Z = (Z_1, ..., Z_d)'$ from $d$ standard normals

3) Set $X = A Z$

4) Return $U = (\Phi(X_1), ..., \Phi(X_d))'$ where $\Phi$ is the standard normal CDF.



```{r simulate truth}

simulate_truth <- function(){
  n_latitude <- 10
  n_longitude <- 10
  d <- n_latitude * n_longitude
  corr <- .4
  
  P <- bandSparse(n = d, k = -c(0,1), diagonals = list(rep(1, d), rep(corr,d)), symm = TRUE)
  A <- t(chol(P))
  Z <- rnorm(n = d) %>% as.matrix()
  X <- A %*% Z
  U <- .2 + pnorm(X[,1]) / 10
  U_matrix <- matrix(U, nrow = n_latitude, ncol = n_longitude, byrow = TRUE)
}

```

```{r plot simulated surface}

```