---
title: "SOC Sampling"
author: "Jake Spertus"
date: "June 9th, 2020"
output: html_notebook
header_includes:
  -\usepackage{amsmath}
  -\usepackage{amsfonts}
  -\usepackage{color}
  -\newcommand{\indep}{\perp \!\!\! \perp}
bibliography: soilcarbonstatistics.bib
---
```{r knitr setup, message = FALSE, warning = FALSE, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r packages, message = FALSE, results = FALSE}
library(tidyverse)
source("sim_functions.R")
```


# Goals



# Setup and Notation

To begin we will consider %SOC as a two-dimensional surface defined over plots $\mathcal{P} \subset \mathbb{R}^2$, where $\mathbb{R}^2$ is a vector space with the usual Euclidean norm $\|u\| = \sqrt{u_1^2 + u_2^2}$ and inner product $\langle u,v \rangle = u_1v_1 + u_2v_2$. Typically, $\mathcal{P} \equiv [a,b]\times[c,d]$, a rectangular plot. For example in [@ryals_impacts_2014] we have a $25 \times 60$ meter plot so that $|a-b| = 25$ and $|c-d| = 60$. A typical element of $\mathcal{P}$ is an ordered pair $(x,y)$, which denotes the position of a point in the plot. Since we typically will consider a single plot, we can define (0,0) to be the lower left hand corner of the plot, so that points $(x,y)$ are relative to the position of the plot itself in $\mathbb{R}^2$. WLOG, we'll typically also take the orientation of the plot to be such that the $y$ direction indicates North/South (latitude) and the $x$ direction indicates East/West (longitude). For example, $(1,2)$ indicates a point 1 meter to the east and 2 meters to the north of the origin. 

Given any point $(x,y)$ there is an associated concentration of soil carbon (%SOC), which we will denote by $z(x,y)$. Note that %SOC is not truly defined at a precise point $(x,y)$ - there is considerable variance at ever smaller scales: a tiny piece of root may be 100\% carbon, while a patch of sand may be 0\% carbon. $z(x,y)$ is better conceptualized as an average over a small window centered at $(x,y)$. In a design-based framework (as we are considering here), $z(x,y)$ is a fixed, unknown number. Typically the parameter of interest is:

$$\mu \equiv \int_\mathcal{P} z(x,y) d\mathcal{P}  = \int_a^b \int_c^d z(x,y)~ dy ~dx$$

This is the "population average" %SOC over the plot $\mathcal{P}$. We wish to estimate $\mu$ using data. 

In field studies, data take the form of finite samples determined at specific values of $x$ and $y$: $\{z(x_1,y_1), z(x_2, y_2), ... z(x_n, y_n) \}$ often we'll denote the data $\{z_1, z_2, ... z_n\} \equiv \{z_i\}_{i=1}^n$ when the locations themselves are not super important. 


# Simple random sampling

### Hoeffding Bound

We can use the *Hoeffding bound* to construct a finite sample confidence interval on the percent carbon estimated using simple random sampling. This bound is finite-sample valid and does not rely on any other assumptions other than the fact that percent carbon is in $[0,1]$. For samples $Z_1, Z_2, ... Z_n$ drawn from a bounded population, the probability that the sample mean is $\epsilon$ far from the population mean is:

$$P(|\bar{Z} - \mathbb{E}[Z_1]| \geq \epsilon) \leq 2\exp\{-2 n \epsilon^2\} $$

If we would like a 95% confidence interval on the sample mean, then we can set the LHS equal to .05 and rearrange:

\begin{align}
.05 &\leq 2 \exp\{- 2 n \epsilon^2\}\\ 
\log(.025) &\leq -2 n \epsilon^2\\
\frac{\log(.025)}{-2\epsilon^2} &\leq n
\end{align}

Thus for a 95\% confidence interval with width $\epsilon$ we would need $\frac{\log(.025)}{-2\epsilon^2}$ samples. This is a lot. Below we plot $n$ as a function of $\epsilon$, the desired precision, on the $\log_{10}$ scale.



```{r}
get_n_hoeffding <- function(epsilon, alpha = .05){
  #compute the hoeffding bound given a level of precision (epsilon) and a level of desired confidence (alpha)
  n <- log(alpha / 2) / (-2 * epsilon^2) 
  n
}
```





### Central Limit Theorem Bound

On the other hand, we could consider the *central limit theorem* and an upper bound on the population variance. Since $Z_i \in [0,1]$, we have that $\sigma^2 \leq 1/4$. The central limit theorem gives an asymptotic distribution of the variation of the sample mean from the population mean:

$$\mathbb{P}(|\bar{Z} - \mu| > \epsilon) = 2 \bigg (1 - \Phi \big (\frac{\epsilon}{\sigma / \sqrt{n}} \big ) \bigg ) \leq 2 \big (1-\Phi ( 4\epsilon \sqrt{n}) \big )$$
Therefore if we wish to bound deviations from the mean by $\epsilon$ to occur with probability less than $\alpha$ we can collect $n$ so that:

\begin{align}
2(1 - \Phi(4 \epsilon \sqrt{n})) &\leq \alpha\\
&\implies \\
\Phi(4 \epsilon \sqrt{n}) \geq 1 - \alpha/2\\
&\implies\\
n \geq \frac{\Phi^{-1}(1-\alpha/2)^2}{16 \epsilon^2}
\end{align}

```{r}
#population variance is 1/4 in the worst case for pct carbon
get_n_clt <- function(epsilon, alpha = .05, variance = 1/4){
  #compute the sample sizes needed using a wald interval
  n <- (qnorm(1 - alpha/2)^2) / (16 * epsilon^2)
  n
}
```


### Comparing Bounds

```{r}
epsilon_grid <- seq(.01, .1, length.out = 100)
n_vs_epsilon <- data.frame(n_hoeffding = get_n_hoeffding(epsilon_grid), n_clt = get_n_clt(epsilon_grid), epsilon = epsilon_grid) %>%
  pivot_longer(cols = c("n_hoeffding","n_clt"))



ggplot(data = n_vs_epsilon, aes(x = epsilon, y = value, colour = name)) +
  geom_line() +
  theme_linedraw() +
  scale_y_log10() +
  labs(y = "n", x = "epsilon")
```



# Stratified Sampling

Stratified sampling should in general be much more efficient than simple random sampling whenever heterogeneity can be reduced by stratification. 

As a preliminary example, consider an extreme case where half of a plot is 0\% SOC and half of a plot is 100\% SOC. This would be like if the Sahara and a peat bog were right next to each other except more extreme (peat is not more than 45\% carbon). The plot $\mathcal{P}$ can be partioned into areas $\mathcal{P}_0 = \{(x_i,y_i) \in \mathcal{P} \mid z(x_i,y_i) = 0\}$ and $\mathcal{P}_1 = \{(x_i,y_i) \in \mathcal{P} \mid z(x_i,y_i) = 1\}$. The variance of a single point drawn by simple random sampling is the population variance:
\begin{align*}
\mathbb{V}[Z_i] &= \int_{\mathcal{P}} (Z_i - \mathbb{E}[Z_i])^2 d\mathcal{P}\\
&= \int_{\mathcal{P}_0} (Z_i - 1/2)^2 d\mathcal{P}_0 + \int_{\mathcal{P}_1} (Z_i - 1/2)^2 d\mathcal{P}_1\\ 
&= \int_{\mathcal{P}_0} 1/4 ~ d\mathcal{P}_0 + \int_{\mathcal{P}_1} 1/4 ~ d\mathcal{P}_1\\
&= 1/8 + 1/8\\ 
&= 1/4
\end{align*}

If these values are uniformly distributed throughout $\mathcal{P}$ there is nothing to be done. But suppose they are distributed so that the right half of the plot is all $Z_i = 1$ and the left half of the plot is all so that $Z_i = 0$. Then the variance within these halves is 0, and a single random sample from each half suffices to completely describe the %SOC there. 

